RegressionPvNP

### 2017-05-05 RegressionPvNP.Rmd
…

```{r setup, include=FALSE}
library(ggplot2)
library(ggfortify)
library(plyr)
library(pander)
library(Hmisc)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)

load('patients.Rdata', verbose=T)

@@

## Patient info file
patient.file = grep('All_patient_info.xls', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

patient.info = read.patient.info(patient.file)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

@@

```{r apclust-func, echo=F, message=F, warning=F}
suppressPackageStartupMessages( library(apcluster) )

@@

apclust<-function(segdata) {
  x1 = apclust.data(segdata)
  # q=0 minimize off-diagonal similarity
  ac = apcluster(negDistMat(r=2), x1, details=T, convits=25, q=0)
  return(list('apres'=ac, 'data'=x1))
}

net.similarity<-function(ac) {
  if (is.list(ac))
    ac = ac$apres
  ac@netsim
}

sum.similarity<-function(ac) {
  if (is.list(ac))
    ac = ac$apres
  ac@dpsim
}

clusters<-function(ac) {
  if (is.list(ac))
    ac = ac$apres
  ac@exemplars
}

plotAC<-function(aclist) {
  plot(aclist$apres, aclist$data)
}

heatmapAC<-function(aclist) {
  heatmap(aclist$apres)
}

# First vs last samples
#x1 = apclust.data(patient.data[[pt]]$seg.vals[,c(1:4, 6, ncol(patient.data[[pt]]$seg.vals))], patient.data[[pt]]$info$Samplename)
```

@@

tileFile = paste(as.character(tile.w),'_tiledpts.Rdata', sep='')
if (file.exists(tileFile)) {
  load(tileFile, verbose=T)
} else {
  grList = lapply(patient.data, function(df) {
    df$info = arrange(df$info, Endoscopy.Year, Pathology)

    x1 = apclust.data(df$seg.vals, df$info$Samplename)
    head(x1)

@@

    pt = unique(df$info$Patient)
    return( makeGRangesFromDataFrame(cbind(segnames,pt, x1), keep.extra.columns = T) )
  })

  sampleNames = unlist(sapply(patient.data, function(df) {
    df$info = arrange(df$info, Endoscopy.Year, Pathology)
    x1 = apclust.data(df$seg.vals, df$info$Samplename)
    colnames(x1)
  }))

  tile.segments<-function(tiles, gr, mergedSegments) {
    ov = findOverlaps(tiles, gr)
    for (qh in unique(queryHits(ov))) {
      print(qh)
      currentTile = tiles[[qh]]
      curov = findOverlaps(currentTile, gr)

@@

  for (gr in grList) {
    print(unique(gr$pt))
    mergedDf = tile.segments(tiles, gr, mergedDf)
  }

@@

  save(mergedDf, file=tileFile)
}

@@

```{r labelsPNP, echo=F, message=F, include=F}
## binomial: dysplasia 1, BE 0

labels = unlist(sapply(patient.data, function(df) {
  df$info = arrange(df$info, Endoscopy.Year, Pathology)
  label = as.integer(df$info$Status == 'P') #as.integer(df$info$Pathology %in% c('HGD', 'IMC'))
  names(label) = df$info$Samplename
  return(label)
}))
names(labels) = sub('.*\\.', '',  names(labels))

@@

With all `r ncol(dysplasia.df)` regions of the genome, first check if/which regression may be appropriate. It appears that pure lasso may provide the best balance for number of features with non-zero coefficients.

```{r cvglmnetPNP, echo=F, warning=F, message=F}
plot.multi.cv.glmnet<-function(x,y,alpha=c(0,1), family="binomial", title='cv.glmnet') {
  plots <- list(  )
  cv.models = data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('class','lambda','cvm'))))
  for (a in alpha) {
    cv = cv.glmnet(x, y, alpha=a, family=family, standardize=F)
    cv.models = rbind(cv.models, cbind('class'=paste('cv',a,sep=''), 'lambda'=log(cv$lambda), 'cvm'=cv$cvm))
    plots[[as.character(a)]] = autoplot(cv, main=paste('alpha',a), ylab=cv$name)
  }
  cv.models[c('lambda','cvm')] = lapply(cv.models[c('lambda','cvm')], function(x) as.numeric(as.character(x)))

@@

## xval by patients

Split into 5ths by patient, fit and predict at each split, select min lambda, get coefficients from the initial model using that value.

```{r xvalpt, message=F, warning=F, echo=F, fig.height=6, fig.width=6}
precisionRecall<-function(actual, predicted) {
    retrieved <- sum(predicted==1)
    recall <- sum(predicted==1 & actual==1) / sum(actual==1)
    precision <- sum(predicted==1 & actual==1) / retrieved
    f1 = signif(2*(1/(1/precision+1/recall)), 2)
    return(data.frame('precision'=precision, 'recall'=recall, 'F1'=signif(f1, 2)))
  }

pi.hat<-function(x) exp(x)/(1+exp(x))

@@

cv.patient.glmnet<-function(x,y,lambda,patients) {
  message(paste("Running", length(unique(patients$group)), "folds"))

  cv.class = matrix(nrow=5, ncol=length(lambda))
  auc = matrix(nrow=5, ncol=length(lambda))

  fit.e = list()
  for (i in 1:length(unique(patients$group))) {  ## This is not stable. 
    message(paste(i, "fold"))
    test.rows = which(rownames(x) %in% subset(patients, group == i)$Samplename)
    test = x[test.rows,]
    training = x[-test.rows,]
    # pre-spec lambda seq
    fit <- glmnet(training, y[-test.rows], lambda=lambda, family='binomial', alpha=1) 
    # autoplot(fit) + theme(legend.position="none")

    pred.class <- predict(fit, test, type='class') # check this later
    auc[i,] = apply(pred.class, 2, function(pd) {
      if (sum(as.numeric(pd)) <= 0) return(0)
      #as.numeric(pROC::auc(pROC::roc(response=as.numeric(pd), predictor=labels[test.rows])))  
      precisionRecall(labels[test.rows], as.numeric(pd))$F1
    })

    # careful: matrix
    pred <- pi.hat(predict(fit, test)) 
    cv.class[i,] = apply(pred, 2, function(p) {
      # Confusion matrix
       #cmat = table(labels[test.rows], p>0.5)
       #sum(diag(cmat))/sum(cmat)
      # quantitiative, pos results + neg results / number of test rows
      (p%*%labels[test.rows] + (1-p) %*% (1-labels[test.rows]))/ length(test.rows)
    })

    fit.e[[i]] = fit
  }

  df = cbind.data.frame('lambda.at'=1:ncol(cv.class),'mean'=colMeans(cv.class), 
                        'sme'=apply(cv.class, 2, sd)/5, 
                        'sd'=apply(cv.class, 2, sd), 
                        'lambda'=lambda)

  lambda.opt = lambda[which.max(df$mean)-1]
  nzcf = lapply(fit.e, non.zero.coef, s=lambda.opt)

  gp = ggplot(df, aes(y=mean,x=lambda.at)) + geom_point() + 
          geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + labs(y='mean of confusion matrix (P)') +
          geom_point(data=subset(df, mean == max(mean)), aes(y=mean, x=lambda.at, colour="max"), size=3 ) +
          geom_point(data=df[which.max(df$mean)+1,], aes(y=mean, x=lambda.at, colour="max+1"), size=3) +
          geom_point(data=df[which.max(df$mean)-1,], aes(y=mean, x=lambda.at, colour="max-1"), size=3) +
          geom_hline(yintercept=df[which.max(df$mean)-1,'mean'], colour='grey') +
          annotate("text", x=5, y=max(df$mean)+0.0005, label=round(df[which.max(df$mean)-1,'mean'],2)) +
          theme(axis.title.x = element_text(size = 15, vjust=-.2)) + scale_colour_discrete(name = "Best lambda values") +
          labs(title="CV Performance")

  return(list('cv.class'=cv.class, 'f1'=auc[which.max(df$mean)-1], 'lambda.opt'=lambda.opt, 'lambdas'=df, 'plot'=gp, 'non.zero.cf'=nzcf))
}

fit0 <- glmnet(dysplasia.df, labels, alpha = 1) # all patients
autoplot(fit0, xvar='lambda', main='fit0, all samples') + theme(legend.position='none') 
lambda <- fit0$lambda

# split patients in 5ths
s = {set.seed(2); sample(rep(seq(5), length = length(unique(pts$Patient))))}
patients = merge(pts, cbind('Patient'=unique(pts$Patient), 'group'=s), by="Patient")

# This just makes sure the sets don't become too unbalanced with regards to the labels.
sets = table(cbind(patients, labels[patients$Samplename])[,c(3:4)])
while (length(which(sets/rowSums(sets) < 0.35)) >= 2 | length(which(sets/rowSums(sets) == 0)) > 0 ) {
  print(".")
  s = sample(rep(seq(5), length = length(unique(pts$Patient))))
  patients = merge(pts, cbind('Patient'=unique(pts$Patient), 'group'=s), by="Patient")
  sets = table(cbind(patients, labels[patients$Samplename])[,c(3:4)])
}

cv.patient = cv.patient.glmnet(x=dysplasia.df, y=labels, lambda, patients)
lambda.opt = cv.patient$lambda.opt

coef.opt = as.data.frame(non.zero.coef(fit0, lambda.opt))

grid.arrange(cv.patient$plot)

## check how often the feature(s) is selected at that lamda in each split. "stability selection"

for (i in 1:length(cv.patient$non.zero.cf)) {
  df = as.data.frame(cv.patient$non.zero.cf[[i]])
  colnames(df) = i
  coef.opt = merge(coef.opt, df, by='row.names', all.x=T)
  rownames(coef.opt) = coef.opt$Row.names
  coef.opt$Row.names = NULL
  coef.opt[,(i+1)] = as.integer(!is.na(coef.opt[,(i+1)]))
}
coef.opt = coef.opt[order(sapply(rownames(coef.opt), function(x) as.numeric(unlist(strsplit(x, ':'))[1])  )),]

@@

`r pander(coef.opt, justify='left', caption='Features from the glm fitted with all data. Each numbered column indicates if the feature was identified in that split.')`

At a the minimum lambda-1se (green dot) the value is `r signif(lambda.opt, 3)`. In the initial model `r length(coef.opt)` coefficients have a non-zero value.

@@

Just as a point, the initial model is tested against `r nrow(dysplasia.df)` samples with `r ncol(dysplasia.df)` features.

```{r noHGD, echo=F, message=F, warning=F}
`%nin%` <- Negate(`%in%`)

@@

samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)

cv.nohgd = cv.patient.glmnet(x=dysplasia.df[samples,], y=labels[samples], lambda, subset(patients, Samplename %in% samples))
#lambda.opt = cv.patient$lambda.opt
grid.arrange(cv.nohgd$plot + ggtitle("CV Performance, No HGD/IMC samples"))
#coef.opt = as.data.frame(non.zero.coef(fit0, lambda.opt))

samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD'))$Samplename)
cv.nolgd = cv.patient.glmnet(x=dysplasia.df[samples,], y=labels[samples], lambda, subset(patients, Samplename %in% samples))
#lambda.opt = cv.patient$lambda.opt
grid.arrange(cv.nolgd$plot + ggtitle("CV Performance, No HGD/IMC/LGD samples"))

@@

>>>>

```{r setup, include=FALSE}
library(ggplot2)
library(ggfortify)
library(plyr)
library(pander)
#library(Hmisc)
library(reshape2)
library(gridExtra)
library(GenomicRanges)
library(glmnet)

load("Test_patients.Rdata", verbose=T)
validation.patient.data = patient.data

dataset = 'Training'
load(paste(dataset, '_patients.Rdata', sep=''), verbose=T)

@@

## Patient info file
patient.file = grep('All_patient_info.txt', data.files, value=T)
if (length(patient.file) != 1)
  stop(paste("Missing/too many patient info file(s) in", data))

patient.info = read.patient.info(patient.file, set=dataset)
patient.info$Patient = gsub("/", "_", patient.info$Patient)
head(patient.info)

validation.patient.info = read.patient.info(patient.file, set='Test')
sum.validation.info = summarise.patient.info(validation.patient.info)

@@

```{r apclust-func, echo=F, message=F, warning=F}
#suppressPackageStartupMessages( library(apcluster) )

@@

# apclust<-function(segdata) {
#   x1 = apclust.data(segdata)
#   # q=0 minimize off-diagonal similarity
#   ac = apcluster(negDistMat(r=2), x1, details=T, convits=25, q=0)
#   return(list('apres'=ac, 'data'=x1))
# }

@@

tile.sample.segments<-function(tileFile, sample.data) {
  if (file.exists(tileFile)) {
    load(tileFile, verbose=T)
  } else {
    grList = lapply(sample.data, function(df) {
      df$info = arrange(df$info, Endoscopy.Year, Pathology)

      x1 = apclust.data(df$seg.vals, df$info$Samplename)

@@

      pt = unique(df$info$Patient)
      return( makeGRangesFromDataFrame(cbind(segnames,pt,x1), keep.extra.columns=T) )
    })

    sampleNames = unlist(sapply(sample.data, function(df) {
      df$info = arrange(df$info, Endoscopy.Year, Pathology)
      x1 = apclust.data(df$seg.vals, df$info$Samplename)
      colnames(x1)
    }))

    tile.segments<-function(tiles, gr, mergedSegments) {
      ov = findOverlaps(tiles, gr)
      for (qh in unique(queryHits(ov))) {
        #print(qh)
        currentTile = tiles[[qh]]
        curov = findOverlaps(currentTile, gr)

@@

    for (gr in grList) {
      #print(unique(gr$pt))
      mergedDf = tile.segments(tiles, gr, mergedDf)
    }

@@

    save(mergedDf, file=tileFile)
  }
  return(mergedDf)
}

mergedDf = tile.sample.segments(paste(as.character(tile.w),'_tiledpts.Rdata', sep=''), patient.data)

#validatioMergedDf = tile.sample.segments(paste(as.character(tile.w),'validation_tiledpts.Rdata', sep=''), validation.patient.data)

@@

```{r labelsPNP, echo=F, message=F, include=F}
## binomial: dysplasia 1, BE 0
labels = unlist(sapply(patient.data, function(df) {
  df$info = arrange(df$info, Endoscopy.Year, Pathology)
  label = as.integer(df$info$Status == 'P') #as.integer(df$info$Pathology %in% c('HGD', 'IMC'))
  names(label) = df$info$Samplename
  return(label)
}))
names(labels) = sub('.*\\.', '',  names(labels))

@@

```{r cvglmnetPNP, echo=F, warning=F, message=F, eval=F}
#With all `r ncol(dysplasia.df)` regions of the genome, first check if/which regression may be appropriate. It appears that pure lasso may provide the best balance for number of features with non-zero coefficients.

plot.multi.cv.glmnet<-function(x,y,alpha=c(0,1), family="binomial", title='cv.glmnet') {
  plots <- list(  )
  cv.models = data.frame(matrix(ncol=3,nrow=0,dimnames=list(c(), c('class','lambda','cvm'))))
  for (a in alpha) {
    cv = cv.glmnet(x, y, alpha=a, family=family, standardize=F)
    cv.models = rbind(cv.models, cbind('class'=paste('cv',a,sep=''), 'lambda'=log(cv$lambda), 'cvm'=cv$cvm))
    plots[[as.character(a)]] = autoplot(cv, main=paste('alpha',a), ylab=cv$name)
  }
  cv.models[c('lambda','cvm')] = lapply(cv.models[c('lambda','cvm')], function(x) as.numeric(as.character(x)))

@@

## Build/Cross-validate by patient

Keeping in mind that the matrix is built on a sample, not patient, basis while the labels (Progressor/Non) are on a per-patient basis I generated the cross validated models for various values of alpha by running 10 or more iterations of 5-fold cross validation that pulled out all the samples for 5 patients at each fold.  1000 lambda values.

```{r cv.funcs, message=F, warning=F, echo=T, fig.height=6, fig.width=6}
pi.hat<-function(x) exp(x)/(1+exp(x))

@@

create.patient.sets<-function(pts, n, splits, minR=0.2) {
  # This function just makes sure the sets don't become too unbalanced with regards to the labels.
  check.sets<-function(df, grpCol, min) {
    sets = table(cbind.data.frame('set'=df[[grpCol]], 'labels'=labels[df$Samplename]))
    while ( (length(which(sets/rowSums(sets) < minR) ) >= 2 | length(which(sets/rowSums(sets) == 0)) > 0) ) {
      #print(sets/rowSums(sets))
      s = sample(rep(seq(5), length = length(unique(df$Patient))))
      df2 = merge(df, cbind('Patient'=unique(df$Patient), 'tmpgrp'=s), by="Patient")
      df[[grpCol]] = df2$tmpgrp
      sets = table(cbind.data.frame('set'=df[[grpCol]], 'labels'=labels[df$Samplename]))
    }
  return(df[[grpCol]])
  }

  s = sample(rep(seq(splits), length = length(unique(pts$Patient))))
  patients = merge(pts, cbind('Patient'=unique(pts$Patient), 'group'=s), by="Patient")  
  colnames(patients)[3] = c('fold.1')
  patients$fold.1 = check.sets(patients, 'fold.1', minR)

  for (i in 2:n) {
    s = sample(rep(seq(5), length = length(unique(pts$Patient))))
    patients = merge(patients, cbind('Patient'=unique(patients$Patient), 'group'=s), by="Patient")  
    foldcol = grep('group',colnames(patients))
    colnames(patients)[foldcol] = paste('fold',i,sep='.')
    patients[[paste('fold',i,sep='.')]] = check.sets(patients, paste('fold',i,sep='.'))
  }
  return(patients)
}

crossvalidate.by.patient<-function(x,y,lambda,pts,a=1,nfolds=10, splits=5, fit=NULL, minR=0.2) {
  message(paste("Running", splits, "splits",nfolds,"times on", paste(dim(x), collapse=':'), 'alpha=',a ))
  fit.e = list()
  pts = create.patient.sets(pts, nfolds, splits, minR)
  cv.pred = (matrix(nrow=0, ncol=length(lambda)))
  for (n in 1:nfolds) {  
    message(paste(n, "fold"))
    setCol = grep(paste('^fold.',n,'$',sep=''), colnames(pts))

    cv.class = matrix(nrow=splits, ncol=length(lambda))
    for (i in 1:splits) { 
      message(paste(i, "split"))
      test.rows = which(rownames(x) %in% pts[which(pts[,setCol] == i), 'Samplename'])
      test = x[test.rows,]
      training = x[-test.rows,]
      # pre-spec lambda seq
      fit <- glmnet(training, y[-test.rows], lambda=lambda, family='binomial', alpha=a) 
      # autoplot(fit) + theme(legend.position="none")
      # careful: matrix
      pred <- pi.hat(predict(fit, test, type='link')) 
      cv.class[i,] = apply(pred, 2, function(p) {
        # Confusion matrix: quantitiative, pos results + neg results / number of test rows
        (p%*%labels[test.rows] + (1-p) %*% (1-labels[test.rows]))/ length(test.rows)
      })

      fit.e[[length(fit.e)+1]] = fit
    }
  cv.pred = rbind(cv.pred, cv.class)    
  }

  df = cbind.data.frame('lambda.at'=1:ncol(cv.pred),
                        'mean'= colMeans(cv.pred), 
                        'sme'= apply(cv.pred, 2, sd)/sqrt(nrow(cv.pred)), 
                        'sd'= apply(cv.pred, 2, sd), 
                        'lambda'= lambda)

  df2 = df
  df2$lambda = log(lambda)
  df2$mean = 1-df2$mean

  df2$lambda.min = df2$mean == min(df2$mean)
  df2$lambda.1se = FALSE
  df2[df2$lambda < subset(df2, mean == min(mean))$lambda-sd(df2$lambda),][1,'lambda.1se'] = TRUE

  lambda.min = lambda[subset(df2, lambda.min == T)$lambda.at]
  lambda.1se = lambda[subset(df2, lambda.1se == T)$lambda.at]
  nzcf = lapply(fit.e, non.zero.coef, s=lambda.min)

  gp = ggplot(df2, aes(y=mean,x=lambda)) + geom_point() + 
    geom_errorbar(aes(ymin=mean-sme, ymax=mean+sme)) + 
    geom_point(data=subset(df2, lambda.min == T), aes(y=mean, x=lambda, colour="lambda.min"), size=2 ) +          
    geom_vline(xintercept = subset(df2, lambda.min == T)$lambda, colour='grey') +
    annotate("text", x=subset(df2, lambda.min == T)$lambda, y=0.35, 
             label=round(df[subset(df2, lambda.min == T)$lambda.at,'mean'],3))

  if (length(lambda.1se) > 0) {
    gp = gp + geom_point(data=subset(df2, lambda.1se == T), aes(y=mean, x=lambda, colour='lambda.1se'), size=2 ) +
      geom_vline(xintercept = subset(df2, lambda.1se == T)$lambda, colour='grey') +
      annotate("text", x=subset(df2, lambda.1se == T)$lambda, y=0.35, 
               label=round(df[subset(df2, lambda.1se == T)$lambda.at,'mean'],3)) 
  }
   gp = gp + theme( legend.position='bottom') + 
          scale_colour_discrete(name = "") +
          labs(title=paste(splits,' splits, ',nfolds,' folds, alpha=',a, sep=''), y='1-mean (P)', x='log(lambda)')

  if (!is.null(fit)) {
    df2$nzcoef = sapply(df$lambda, function(l) length(non.zero.coef(fit, l)))

    coef.min = subset(df2, lambda.min == T)$nzcoef
    coef.1se = subset(df2, lambda.1se == T)$nzcoef
    d = coef.1se-coef.min

    coef.text = arrange(subset(df2,nzcoef %in% c(0,  coef.min, coef.1se, max(nzcoef) )), -lambda.min, -lambda.1se)
    coef.text = arrange(coef.text[!duplicated(coef.text$nzcoef),], -nzcoef)

    gp = gp + annotate("text", y=max(df2$mean), x=coef.text$lambda, label=coef.text$nzcoef)
  }

  return(list('max.cm'=df[which.max(df$mean)-1,'mean'], 
              'lambda.min'=lambda.min, 'lambda.1se'=lambda.1se, 'lambdas'=df, 'plot'=gp, 'non.zero.cf'=nzcf))
}

coef.stability<-function(opt, nz.list) {
  for (i in 1:length(nz.list)) {
    df = as.data.frame(nz.list[[i]])
    colnames(df) = i
    opt = merge(opt, df, by='row.names', all.x=T)
    rownames(opt) = opt$Row.names
    opt$Row.names = NULL
    opt[,(i+1)] = as.integer(!is.na(opt[,(i+1)]))
  }
  opt = opt[order(sapply(rownames(opt), function(x) as.numeric(unlist(strsplit(x, ':'))[1])  )),]
  return(opt)
}
```

```{r xvalpt, message=F, warning=F, echo=T, fig.height=12, fig.width=12}
plots = list()
coefs = list()
folds = 10; splits=5
alpha.values = c(0,0.5,0.7,0.8,0.9,1)
for (a in alpha.values) {
  fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=1000, family='binomial') # all patients
  autoplot(fit0, xvar='lambda', main=paste('fit0, all samples, alpha=',a,sep='')) + theme(legend.position='none') 
  lambda0 <- fit0$lambda

  cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=lambda0, pts=pts, a=a, nfolds=folds, splits=splits, fit=fit0)

  coef.1se = as.data.frame(non.zero.coef(fit0, cv.patient$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.patient$non.zero.cf)

  plots[[as.character(a)]] = cv.patient$plot + theme(legend.position='bottom') + ggtitle(paste('alpha=',a,sep=''))
}

do.call(grid.arrange, c(plots, top='All samples, 10fold, 5 splits', ncol=2))

## check how often the feature(s) is selected at that lamda in each split. "stability selection"
coef.stable = lapply( coefs[c('0.8','0.9','1')], function(cf) {
  sort(rowSums(cf[,-1]), decreasing=T)
})

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=3, 
                dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))

for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1

@@

Values that are closer to full lasso (alpha=1) are pretty similar, however at full lasso we see fewer features than if we regularize it at 0.8.

### Feature stability

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

They also share all of the non-zero coefficients (0.8 identifies the most non-zero coefs `r max(sapply(coef.stable, length))`).

`r pander(cfs, justify='left')`

#### Did I understand correctly?

In order to predict, my trained model is then my original model, but using the optimal lambda value?
This isn't quite correct because the model was trained including these patients, just an example.

```{r}
fit0 <- glmnet(dysplasia.df, labels, alpha=a, nlambda=200, family='binomial') # all patients
cv.patient = crossvalidate.by.patient(x=dysplasia.df, y=labels, lambda=fit0$lambda, pts=pts, a=1, nfolds=20, splits=5, fit=fit0)

# Progressor, minus the HGD sample(s)
cbind.data.frame(
  predict(fit0, dysplasia.df[patient.data$PR1_HIN_042$info$Samplename[-(9:10)],], s=cv.patient$lambda.1se, type='response'),
  predict(fit0, dysplasia.df[patient.data$PR1_HIN_042$info$Samplename[-(9:10)],], s=cv.patient$lambda.1se, type='class'))

# Non-P, all correctly predicted
cbind.data.frame(predict(fit0, dysplasia.df[patient.data$AD0610$info$Samplename,], s=cv.patient$lambda.1se, type='response'),
                 predict(fit0, dysplasia.df[patient.data$AD0610$info$Samplename,], s=cv.patient$lambda.1se, type='class'))

@@

Just as a point, the initial model is trained against `r nrow(dysplasia.df)` samples with `r ncol(dysplasia.df)` features.  What happens if we retrain it without HGD?

```{r noHGD, echo=T, message=F, warning=F, fig.height=8, fig.width=8}
`%nin%` <- Negate(`%in%`)

@@

# No HGD/IMC
samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC'))$Samplename)

plots = list()
coefs = list()
for (a in alpha.values) {
  fit1 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit1, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda1 <- fit1$lambda

  cv.nohgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda1, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit1)

  plots[[as.character(a)]] = cv.nohgd$plot + ggtitle(paste('alpha=',a,sep=''))
  coef.1se = as.data.frame(non.zero.coef(fit1, cv.nohgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nohgd$non.zero.cf)
}

do.call(grid.arrange, c(plots, top="No HGD/IMC samples", ncol=2))

## check how often the feature(s) is selected at that lamda in each split. "stability selection"
coef.stable = lapply( coefs[c('0.8','0.9','1')], function(cf) {
  sort(rowSums(cf[,-1]), decreasing=T)
})

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=3, 
                dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))

for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1

#### Feature Stability

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

They also share all of the non-zero coefficients (`r which.max(sapply(coef.stable, length))` identifies the most non-zero coefs `r max(sapply(coef.stable, length))`).

`r pander(cfs, justify='left')`

### Remove HGD/IMC & LGD Samples

Now remove all LGD samples from the progressor's and retrain.

```{r noLGD, echo=T, message=F, warning=F, fig.height=8, fig.width=8}
# No LGD
samples = intersect(rownames(dysplasia.df), c(subset(info, Status == 'NP')$Samplename, subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD') & Status == 'P')$Samplename))

# No HGD/IMC/LGD in all patients
samples = intersect(rownames(dysplasia.df), subset(info, Pathology %nin% c('HGD', 'IMC', 'LGD'))$Samplename)

plots = list()
coefs = list()
for (a in alpha.values) {
  fit2 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=a, family='binomial', nlambda=1000) # all patients
  autoplot(fit2, xvar='lambda', main='fit1, no HGD/IMC samples') + theme(legend.position='none') 
  lambda2 <- fit2$lambda

  cv.nolgd = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=lambda2, pts=subset(pts, Samplename %in% samples), a=a, nfolds=folds, splits=splits, fit=fit2)

  plots[[as.character(a)]] = cv.nolgd$plot + ggtitle(paste('alpha=',a,sep=''))

  coef.1se = as.data.frame(non.zero.coef(fit2, cv.nolgd$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.nolgd$non.zero.cf)
}

do.call(grid.arrange, c(plots, top="No HGD/IMC/LGD samples", ncol=2))

## check how often the feature(s) is selected at that lamda in each split. "stability selection"
coef.stable = lapply( coefs[c('0.8','0.9','1')], function(cf) {
  sort(rowSums(cf[,-1]), decreasing=T)
})

cfs = as.data.frame(matrix(data=0,nrow=length(unique(names(table(unlist(lapply(coef.stable, names)))))), ncol=3, 
                dimnames=list( unique(names(table(unlist(lapply(coef.stable, names))))), names(coef.stable) )))

for (i in names(coef.stable)) 
  cfs[intersect(rownames(cfs), names(coef.stable[[i]])), i] = 1

#### Feature Stability

`r pander(as.data.frame(coef.stable[['1']])/(folds*splits), justify='left', caption='Percentage of features from the glm fitted with all data. Numbers indicate the total number of folds that feature was identified in.')`

They also share all of the non-zero coefficients (`r which.max(sapply(coef.stable, length))` identifies the most non-zero coefs `r max(sapply(coef.stable, length))`).

`r pander(cfs, justify='left')`

### Leave one patient out

Using the 'all sample' fit, run a LOO (patient)

```{r loop, echo=T, message=F, warning=F, fig.height=6, fig.width=6, eval=F}
fit0 <- glmnet(dysplasia.df, labels, alpha=1, nlambda=200, family='binomial') # all patients

# Randomize patients
pt.samp = sample(sum.patient.data$Patient)
performance.at.1se = c()
coefs = list()
for (pt in pt.samp) {
  samples = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Patient != pt)$Patient], function(pt) pt$info$Samplename)))

  cv.loo = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts= subset(pts, Samplename %in% samples), a=1, nfolds=20, splits=5, fit=fit0)

  performance.at.1se = c(performance.at.1se, subset(cv.loo$lambdas, lambda == cv.loo$lambda.1se)$mean)

  coef.1se = as.data.frame(non.zero.coef(fit0, cv.loo$lambda.1se))
  coefs[[as.character(a)]] = coef.stability(coef.1se, cv.loo$non.zero.cf)
}

ggplot(as.data.frame(performance.at.1se), aes(y=performance.at.1se, x='LOO')) + ylim(0.6,0.8) +
  geom_boxplot( fill='lightblue', color='darkgrey', outlier.fill=NA, outlier.color=NA) + geom_jitter() +
    labs(y='performance at lambda.1se', x='', title='Performance for Leave One (patient) Out, at alpha=1')

## Compare progressors with many samples vs few

```{r, echo=T, message=F, warning=F, eval=F}
m = median(subset(sum.patient.data, Status == 'P')$total.samples)

nonps = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'NP')$Patient], function(df) df$info$Samplename)))

# Few
samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples < m)$Patient], function(pt) pt$info$Samplename))))

fit0 <- glmnet(dysplasia.df[samples,], labels[samples], alpha=1, nlambda=200, family='binomial') # all patients
autoplot(fit0) + theme(legend.position = 'none')

cv.back.few = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda=fit0$lambda, pts=subset(pts, Samplename %in% samples), splits=5, nfolds=20, a=1, fit=fit0)

# Many
samples = c(nonps, as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Status == 'P' & total.samples >= m)$Patient], function(pt) pt$info$Samplename))))
cv.back.many = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda0, subset(patients, Samplename %in% samples))

grid.arrange(cv.back.few$plot + ggtitle(paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples < m))," Progressors with < median # samples")), 
             cv.back.many$plot + ggtitle(paste(nrow(subset(sum.patient.data, Status == 'P' & total.samples >= m)), "Progressors with >= median # samples")), top=paste("Median number of samples:", m))

## Peel back timepoints

So here, do I use the initial fit (all samples), or should I be using the noHGD fit...

```{r timepoints, echo=T, message=F, warning=F, eval=F}

@@

## Example

This can't be right...

```{r , echo=T, message=F, warning=F, eval=F}

# prog.samp = sample(subset(sum.patient.data, Status == 'P')$Patient)
# 
# pt = 'AHM0952' # P
# #pt = 'AHM0254' # NP
# testSamples = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Patient == pt)$Patient], function(pt) pt$info$Samplename)))
# 
# #samples = as.vector(unlist(lapply(patient.data[subset(sum.patient.data, Patient != pt)$Patient], function(pt) pt$info$Samplename)))
# #cv.back = crossvalidate.by.patient(x=dysplasia.df[samples,], y=labels[samples], lambda0, subset(patients, Samplename %in% samples))
# #cv.back$plot
# 
# predict(fit0, dysplasia.df[testSamples,], s=lambda.opt, type='response')
# predict(fit0, dysplasia.df[testSamples,], s=lambda.opt, type='class')
# 
# 
# # Random selection of samples
# randomSamples = sample(as.vector(unlist(lapply(patient.data, function(pt) pt$info$Samplename))), 30)
# predict(fit0, dysplasia.df[randomSamples,], s=lambda.opt, type='response')
# 
# cbind(predict(fit0, dysplasia.df[randomSamples,], s=lambda.opt, type='class'),labels[randomSamples])
